Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-01-23 paper_comparison/paper_comparison/paper_comparison.tex
Fix the plot size
fix plot caption

\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of a matrix,
\(\boldsymbol{X}_a = (\boldsymbol{1}, \boldsymbol{X})\) is the matrix of
regressors including the intercept used in the regression equation,
\(\sigma^2\) is the assumed variance of the error term when \(H_0\) is
true, \(\boldsymbol{V}\) is the actual variance of the error term, and
\(\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of the residuals with \(\boldsymbol{Z}\) be any
variables leave out by the model and \(\boldsymbol{\beta}_z\) be the
corresponding coefficients.

To study various factors contributing to the power of the visual test,
the same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. These include the
distribution of the regressor and the type of the simulated model
(modify this according to the result).

@@

\caption{Variations in fitted values, that might affect perception of residual plots. Four different distribution are used.}\label{fig:different-dist}
\end{figure}

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups
used in experiment I. This lineup is produced by the non-linearity model
under \(j = 6\) and fitted values follow a uniform distribution. The
data plot location is \(2^3 - 4\). All five subjects correctly identify
the data plot from this lineup.

@@

\caption{Three different values of $n$ are used in experiment I, II and III to control the strength of the signal.}\label{fig:different-n}
\end{figure}

@@

\caption{Four different values of $\sigma$ are used in the experiment I to control the strength of the signal.}\label{fig:different-sigma}
\end{figure}

@@

\caption{Lineup poly-24 in experiment I. Can you spot the most different plot? \label{fig:example-poly-lineup}}\label{fig:example-poly-lineup}
\end{figure}

@@

\caption{Lineup heter-169 in experiment II. Can you spot the most different plot? \label{fig:example-heter-lineup}}\label{fig:example-heter-lineup}
\end{figure}

@@

All lineups consist of only null plots are planned to be evaluated by 20
subjects. However, presenting only these lineups to subjects are
considered to be bad practices as subjects will lose interest quickly.
Therefore, we plan to collect 6 more evaluations on the 279 lineups with
fitted values follow a uniform distribution, result in
\((36 \times 20 + (4 \times 4 \times 3 + 3 \times 5 \times 3) \times 3 \times 6) / (20-2) = 133\)
subjects recruited for experiment III.

@@

Figure \ref{fig:polypower} shows the estimated power of visual test on
lineups produced by the non-linearity model with fitted values follow a
uniform distribution, against the natural logarithm of the effect
\(log_e(E)\), with a 5\% significance level. At the bottom of the figure
\ref{fig:polypower}, there are a sequence of example residual plots with
increasing levels of \(log_e(E)\). Readers can evaluate them from left
to right and determine at which level the departure from a good residual
plot becomes detectable.

@@

For the non-linearity model, the power curve of RESET test climbs
aggressively from 15\% to 69\% as \(log_e(E)\) increases from 0 to 2,
while power of other tests respond inactively to the change of effect,
showing that RESET test is way more sensitive to the type of model
defects that being considered. Meanwhile, no noticeable visual features
can be spotted from the example residual plots.

In terms of the heteroskedasticity model, the power of Breusch-Pagan
test is also almost always greater than the power of visual test. For
\(0 \leq log_e(E) \leq 2\), where the power curve of the visual test
remains at a low level, the Breusch-Pagan test still have a decent
amount of chance of rejecting \(H_0\). Similarly, the visual feature is
nearly unobservable from the example residual plots.

The power of visual test arises steadily as \(log_e(E)\) increases from
2 to 5 for both non-linearity model and heteroskedasticity model,
suggesting that the effect starts to make significant impact on the
degree of the presence of the designed visual features. This can also be
observed from the example residual plots that when \(log_e(E) = 2.5\), a
weak ``S-shape'' and a weak ``triangle'' shape are presented in Figure
\ref{fig:polypower} and Figure \ref{fig:heterpower} respectively. The
visual pattern becomes much clearer as \(log_e(E)\) increases. At
\(log_e(E) \approx 6\), the power reaches almost 100\%.

@@

Figure \ref{fig:p-value-comparison} provides a mosaic plot showing the
rejection rate of visual tests and conventional tests for both the
non-linearity model and the heteroskedasticity model.

For lineups produced by the non-linearity model, conventional tests
reject 69\% and visual tests reject 32\% of the time. Of the tests
rejected by the conventional test, 46\% are rejected by the visual test,
that is, approximately half as many as the conventional test. There are
no lineups that are rejected by the visual test but not by the
conventional test.

For lineups produced by the non-linearity model, conventional tests
reject 76\% and visual tests reject 56\% of the time. Of the tests
rejected by the conventional test, 73\% are rejected by the visual test,
that is, approximately half as many as the conventional test.

@@

\caption{The single heteroskedasticity lineup that is rejected by the visual test but not by the BP test. The data plot at panel 17 contains a "Butterfly" shape. It has effect size$ = 4.02$, somewhat surprising that it is not detected by the BP test.}\label{fig:heter-example}
\end{figure}

\hypertarget{effect-of-other-factors-used-in-the-simulated-model}{%
\subsection{Effect of other factors used in the simulated
model}\label{effect-of-other-factors-used-in-the-simulated-model}}

@@

For the non-linearity model, we do not observe significant difference in
power of visual test between distributions, except for the discrete
uniform distribution. It is possibility due to the fact that humans have
difficulties in recognising visual patterns from discreteness, which
makes the shape disconnected and incomplete (ref?). In terms of
conventional tests, the discrete uniform distribution is instead the one
with the greatest power, followed by the lognormal distribution, the
uniform distribution and the normal distribution.

For the heteroskedasticity model, the power of visual test on lineups
with uniform distribution has the greatest power. This is as expected
since other three distributions could reduce the chance of revealing the
underlying visual pattern because of uneven data points. Although the
power of visual test under the discrete uniform distribution is the
lowest in the case of non-linearity model, it has a relatively great
power this time. Considering for heteroskedasticity, the visual patterns
are usually detected by connecting the maximum and minimum residuals
separately at different fitted values. It will not be greatly affected
by the discrete uniform distribution compared to the uniform
distribution. Visual tests have low power on lineups with the normal
distribution and the lognormal distribution. This suggests the use of
transformation of fitted values in reading residual plots. In the case
of conventional tests, the discrete uniform distribution consistently
doing great, followed by the uniform distribution, the normal
distribution and the lognormal distribution.

In regards of the type of the simulated model, the power of visual tests
under different distributions of regressor are always lower than the
corresponding conventional tests, which is consistent with the result
from the previous sections.

@@

\caption{Power of conventional tests and visual tests on lineups produced by the non-linearity model under different fitted value distributions.}\label{fig:different-x-dist-poly-power}
\end{figure}

>>>>

\noindent where \(diag(.)\) is the diagonal matrix constructed from the
diagonal elements of a matrix,
\(\boldsymbol{X}_a = (\boldsymbol{1}, \boldsymbol{X})\) is the matrix of
regressors including the intercept used in the regression equation,
\(\sigma^2\boldsymbol{I}\) is the assumed variance of the error term
when \(H_0\) is true, \(\boldsymbol{V}\) is the actual variance of the
error term, and
\(\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of the residuals with \(\boldsymbol{Z}\) be any
variables leave out by the model and \(\boldsymbol{\beta}_z\) be the
corresponding coefficients.

To study various factors contributing to the power of the visual test,
the same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. These include the
distribution of the regressor and the type of the simulation model
(modify this according to the result).

@@

\caption{Variations in fitted values, that might affect perception of residual plots. Four different distributions are used.}\label{fig:different-dist}
\end{figure}

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups
used in experiment I. This lineup is produced by the non-linearity model
with \(j = 6\) and uniform fitted values. The data plot location is
\(2^3 - 4\). All five subjects correctly identify the data plot from
this lineup.

@@

\caption{Number of observations $n$ have an impact on the strength of the signal. Three different values of $n$ are used in experiment I, II and III.}\label{fig:different-n}
\end{figure}

@@

\caption{For the non-lienearity model, the standard deviation of the error term $\sigma$ affects the strength of the signal. Four different values of $\sigma$ are used in the experiment I.}\label{fig:different-sigma}
\end{figure}

@@

\caption{One of the lineups containing non-linearity patterns used in experiment I. Can you spot the most different plot? The data plot is positioned at $2^3 + 1$.}\label{fig:example-poly-lineup}
\end{figure}

@@

\caption{One of the lineups containing heteroskedasticity pattern used in experiment II. Can you spot the most different plot? The data plot is positioned at $3^3 - 3^2$}\label{fig:example-heter-lineup}
\end{figure}

@@

All lineups consist of only null plots are planned to be evaluated by 20
subjects. However, presenting only these lineups to subjects are
considered to be bad practices as subjects will lose interest quickly.
Therefore, we plan to collect 6 more evaluations on the 279 lineups with
uniform fitted values, result in
\((36 \times 20 + (4 \times 4 \times 3 + 3 \times 5 \times 3) \times 3 \times 6) / (20-2) = 133\)
subjects recruited for experiment III.

@@

Figure \ref{fig:polypower} shows the estimated power of visual test on
lineups produced by the non-linearity model with uniform fitted values,
against the natural logarithm of the effect \(log_e(E)\), with a 5\%
significance level. At the bottom of the figure \ref{fig:polypower},
there are a sequence of example residual plots with increasing levels of
\(log_e(E)\). Readers can evaluate them from left to right and determine
at which level the departure from a good residual plot becomes
detectable.

@@

For non-linearity patterns, the power curve of RESET test climbs
aggressively from 15\% to 69\% as \(log_e(E)\) increases from 0 to 2,
while power of other tests respond inactively to the change of effect,
showing that RESET test is way more sensitive to the type of model
defects that being considered. Meanwhile, no noticeable visual features
can be spotted from the example residual plots.

In terms of heteroskedasticity patterns, the power of Breusch-Pagan test
is also almost always greater than the power of visual test. For
\(0 \leq log_e(E) \leq 2\), where the power curve of the visual test
remains at a low level, the Breusch-Pagan test still have a decent
amount of chance of rejecting \(H_0\). Similarly, the visual feature is
nearly unobservable from the example residual plots.

The power of visual test arises steadily as \(log_e(E)\) increases from
2 to 5 for both non-linearity patterns and heteroskedasticity patterns,
suggesting that the effect starts to make significant impact on the
degree of the presence of the designed visual features. This can also be
observed from the example residual plots that when \(log_e(E) = 2.5\), a
weak ``S-shape'' and a weak ``triangle'' shape are presented in Figure
\ref{fig:polypower} and Figure \ref{fig:heterpower} respectively. The
visual pattern becomes much clearer as \(log_e(E)\) increases. At
\(log_e(E) \approx 6\), the power reaches almost 100\%.

@@

Figure \ref{fig:p-value-comparison} provides a mosaic plot showing the
rejection rate of visual tests and conventional tests for both
non-linearity patterns and heteroskedasticity patterns.

For lineups containing non-linearity patterns, conventional tests reject
69\% and visual tests reject 32\% of the time. Of the tests rejected by
the conventional test, 46\% are rejected by the visual test, that is,
approximately half as many as the conventional test. There are no
lineups that are rejected by the visual test but not by the conventional
test.

For lineups containing non-linearity patterns, conventional tests reject
76\% and visual tests reject 56\% of the time. Of the tests rejected by
the conventional test, 73\% are rejected by the visual test, that is,
approximately half as many as the conventional test.

@@

\caption{The single heteroskedasticity lineup that is rejected by the visual test but not by the BP test. The data plot at panel 17 contains a "Butterfly" shape. It has effect size $ = 4.02$, somewhat surprising that it is not detected by the BP test.}\label{fig:heter-example}
\end{figure}

\hypertarget{effect-of-other-factors-used-in-the-simulation-model}{%
\subsection{Effect of other factors used in the simulation
model}\label{effect-of-other-factors-used-in-the-simulation-model}}

@@

For non-linearity patterns, we do not observe significant difference in
power of visual test between distributions, except for the discrete
uniform distribution. It is possibility due to the fact that humans have
difficulties in recognising visual patterns from discreteness, which
makes the shape disconnected and incomplete (ref?). In terms of
conventional tests, the discrete uniform distribution is instead the one
with the greatest power, followed by the lognormal distribution, the
uniform distribution and the normal distribution.

For heteroskedasticity patterns, the power of visual test on lineups
with uniform distribution has the greatest power. This is as expected
since other three distributions could reduce the chance of revealing the
underlying visual pattern because of uneven data points. Although the
power of visual test under the discrete uniform distribution is the
lowest in the case of non-linearity patterns, it has a relatively great
power this time. Considering for heteroskedasticity, the visual patterns
are usually detected by connecting the maximum and minimum residuals
separately at different fitted values. It will not be greatly affected
by the discrete uniform distribution compared to the uniform
distribution. Visual tests have low power on lineups with the normal
distribution and the lognormal distribution. This suggests the use of
transformation of fitted values in reading residual plots. In the case
of conventional tests, the discrete uniform distribution consistently
doing great, followed by the uniform distribution, the normal
distribution and the lognormal distribution.

In regards of the type of the simulation model, the power of visual
tests under different fitted value distributions are always lower than
the corresponding conventional tests, which is consistent with the
result reported in the previous sections.

@@

\caption{Comparison of power of conventional tests and visual tests on lineups produced by the non-linearity model under different fitted value distributions.}\label{fig:different-x-dist-poly-power}
\end{figure}
